---
title: "Gas Stuff"
output: pdf_document
---

## Intro

I want to use historical gas prices so that I can simulate customer behaviour and
test what strategy leads to greatest savings.

## Data

Found historical gas prices (CAN c/L) from the Edmonton (AB) area at
[edmontongasprices.com](http://www.edmontongasprices.com/retail_price_chart.aspx).

Digitized the 1-year chart (2016-09-25, 2017-09-25) using
[Plot Digitizer, 2.6.8](http://plotdigitizer.sourceforge.net/).
Used linear interpolation to get daily prices in CAN/L.

```{r data,echo=FALSE}
x <- read.csv("~/repos/notes/fun/gas-prices.csv")
fun <- approxfun(x$days, x$gasprice, method = "linear")
z <- data.frame(day=1:365, price=fun(1:365)/100)
n <- nrow(z)
summary(z)
plot(z, type="l")
```

## Strategies

Define some constants:

* Tank volume in L: `vol`
* Mileage in L/100km: `mil`
* Avg distance per day: `dis`

```{r const}
vol <- 60
mil <- 9.6
dis <- 50
LperD <- mil*dis/100 # average L/day
```

### Scenario 1: Go until empty

Refill only when low gas indicator light comes on (we ignore the few litres there, but one can adjust tank volume accordingly)


```{r scen1}
z1 <- z # copy for scenario 1
z1$refill <- FALSE # indicates decision to refill
z1$refill[1] <- TRUE
z1$distance <- 0
z1$level <- vol # gas in tank in L
z1$cost <- 0
for (i in 2:n) {
    z1$distance[i] <- z1$distance[i-1] + dis
    doRefill <- z1$level[i-1] - LperD < 0
    if (doRefill) { # refill
        z1$level[i] <- z1$level[i-1] - LperD
        z1$refill[i] <- TRUE
        z1$cost[i] <- ceiling(100 * z1$price[i] * (vol - z1$level[i]))/100
        z1$level[i] <- vol
        z1$distance[i] <- 0
    } else { # not
        z1$level[i] <- z1$level[i-1] - LperD        
    }
}
summary(z1[,-(1:2)])
sum(z1$cost)
plot(z, type="l")
points(z1[z1$refill,1:2], col=2, pch=19)
```

In Scenario 1, we spent `r round(sum(z1$cost),2)` CAD in a year and stopped at the gas station `r sum(z1$refill)` times. We will use this scenario as the baseline.

### Scenario 2: Fill up when cheaper, full when empty

Top up the tank if price is lower, keep driving until empty when price is higher.

```{r scen2v1}
z2 <- z # copy for scenario 2
z2$refill <- FALSE # indicates decision to refill
z2$refill[1] <- TRUE
z2$distance <- 0
z2$level <- vol # gas in tank in L
z2$cost <- 0
for (i in 2:n) {
    z2$distance[i] <- z2$distance[i-1] + dis
    doRefill1 <- z2$level[i-1] - LperD < 0
    doRefill2 <- z2$price[i-1] > z2$price[i]
    if (doRefill1 || doRefill2) { # refill
        z2$level[i] <- z2$level[i-1] - LperD 
        z2$refill[i] <- TRUE
        z2$cost[i] <- ceiling(100 * z2$price[i] * (vol - z2$level[i]))/100
        z2$level[i] <- vol
        z2$distance[i] <- 0
    } else { # not
        z2$level[i] <- z2$level[i-1] - LperD        
    }
}
summary(z2[,-(1:2)])
sum(z2$cost)
sum(z2$cost) - sum(z1$cost)
plot(z, type="l")
points(z2[z2$refill,1:2], col=4, pch=19)
points(z1[z1$refill,1:2], col=2, pch=19)
legend("topleft", bty="n", pch=19, col=c(2,4), 
    legend=paste("Scenario", 1:2))
```

In Scenario 2, we spent `r round(sum(z2$cost),2)` CAD in a year,
saved `r round(sum(z2$cost) - sum(z1$cost), 2)` CAD relative to Scenario 1. For this we had to
stop at the gas station `r sum(z2$refill)` times,
which is `r round(sum(z2$refill)/sum(z1$refill), 1)` times more than under Scenario 1.

### Scenario 2 variations: Don't want to fill up every day

Micro fill-ups can save money, but it is boring to check the prices every day and stop at the gas station. Is there a threshold in terms of number of days since fill-up (which relates to amount of gas missing from the tank under average calculations)?

```{r scen2v2}
fun_limit <- function(Limit=1) {
    z$refill <- FALSE # indicates decision to refill
    z$refill[1] <- TRUE
    z$distance <- 0
    z$level <- vol # gas in tank in L
    z$cost <- 0
    z$dsf <- 0
    for (i in 2:n) {
        z$distance[i] <- z$distance[i-1] + dis
        doRefill1 <- z$level[i-1] - LperD < 0
        z$dsf[i] <- z$day[i] - z$day[rev(which(z$refill))[1]]
        doRefill2 <- z$price[i-1] > z$price[i] && z$dsf[i] > Limit
        if (doRefill1 || doRefill2) { # refill
            z$level[i] <- z$level[i-1] - LperD
            z$refill[i] <- TRUE
            z$cost[i] <- z$price[i] * (vol - z$level[i])
            z$level[i] <- vol
            z$distance[i] <- 0
        } else { # not
            z$level[i] <- z$level[i-1] - LperD        
        }
    }
    z
}
Limit <- 0:14
res <- lapply(Limit, fun_limit)
Savings <- sapply(res, function(z) sum(z$cost) - sum(z1$cost))
Times <- sapply(res, function(z) sum(z$refill) / sum(z1$refill))
plot(Limit, Savings, type="l")
abline(h=0, lty=2)
plot(Limit, Times, type="l")
abline(h=1, lty=2)
```

It looks like the micro-filling strategy pays off, whereas
not sticking to the rule of **fill up when it's cheaper** 
leads to much smaller savings.


### Scenario 3: Fill up when cheaper, partially when expensive and empty

Top up the tank if price is lower, keep driving until empty when price is higher but don't fill up fully.

```{r scen3}
z3 <- z # copy for scenario 2
z3$refill <- FALSE # indicates decision to refill
z3$refill[1] <- TRUE
z3$distance <- 0
z3$level <- vol # gas in tank in L
z3$cost <- 0
p <- 14 # fill up by p*LperD L when expensive
for (i in 2:n) {
    z3$distance[i] <- z3$distance[i-1] + dis
    doRefill1 <- z3$level[i-1] - LperD < 0
    doRefill2 <- z3$price[i-1] > z3$price[i]
    z3$level[i] <- z3$level[i-1] - LperD
    if (doRefill2) { # refill when cheaper
        z3$refill[i] <- TRUE
        z3$cost[i] <- ceiling(100 * z3$price[i] * (vol - z3$level[i]))/100
        z3$level[i] <- vol
        z3$distance[i] <- 0
    } else { # not cheaper
        if (doRefill1) {
            z3$refill[i] <- TRUE
            Fill <- min(vol-z3$level[i], p*LperD)
            z3$cost[i] <- ceiling(100 * z3$price[i] * Fill)/100
            z3$level[i] <- z3$level[i] + Fill
            z3$distance[i] <- 0
        }
    }
}
summary(z3[,-(1:2)])
sum(z3$cost)
sum(z3$cost) - sum(z1$cost)
plot(z, type="l")
points(z3[z3$refill,1:2], col=4, pch=19)
points(z1[z1$refill,1:2], col=2, pch=19)
legend("topleft", bty="n", pch=19, col=c(2,4), 
    legend=paste("Scenario", 1:2))
```

In Scenario 2, we spent `r round(sum(z2$cost),2)` CAD in a year,
saved `r round(sum(z2$cost) - sum(z1$cost), 2)` CAD relative to Scenario 1. For this we had to
stop at the gas station `r sum(z2$refill)` times,
which is `r round(sum(z2$refill)/sum(z1$refill), 1)` times more than under Scenario 1.

